Acceleration of bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-12 trunk/LateX/paper8.tex
Added a new reference and changed a few minor things.

In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time in half.
Our specific contributions are
\begin{enumerate}
\item
\item
\item
\end{enumerate}

@@

\begin{algorithm}[h]
\caption{Pair-symmetric Bilateral Filtering Algorithm}
\label{alg:Example-Algorithm-float-1}
\begin{algorithmic}
\STATE /* D is the set of dependency arrays */
\FORALL{element $s_{i,j}$ in source image s}
\STATE{$D_1[i,j] \leftarrow 0$}
\STATE{$D_2[i,j] \leftarrow 0$}
\ENDFOR
\FORALL{element $s_{i,j}$ in source image $s$}
\STATE{$X_1 \leftarrow 0$}
\STATE{$X_2 \leftarrow 0$}
\FOR{$k=1$ \TO $filter\_hw$}
\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
\ENDFOR
\STATE{syncthreads();}
\ENDFOR
\FOR{$l = -filter\_hw$ \TO $0$}
\STATE{$temp \leftarrow e^{C_{1} l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
\ENDFOR
\STATE $D_1[i,j] \leftarrow D_1[i,j] + X_1$
\STATE $D_2[i,j] \leftarrow D_2[i,j] + X_2$
\ENDFOR
\FORALL{element $s_{i,j}$ in source image s}
\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{table*}[t!]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache/Memory}  & Compiler\\
\hline 
Intel Xeon E5410  & 2.33 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1\\
\hline 
Intel Xeon X7560  & 2.27 $GH_z$  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1\\
\hline 
Intel Core i7-870  & 2.93 $GH_z$  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2\\
\hline 
AMD Opteron 2376  & 2.3 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1\\
\hline 
AMD Opteron 8350  & 2.0 $GH_z$  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 \\
\hline 
AMD Phenom II X6 1045T  & 2.7 $GH_z$  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2 \\
\hline 
NVidia GeForce GTX 280 & 1296 MHz & 240 & 1 & 240 & - & 1 GB GDDR3 & CUDA 2.0\\
\hline
\end{tabular}
\caption{Architectural details of multicore chips employed for experiments}
\label{tab:archs} 
\end{table*}

@@

\subsubsection{Input Image}
In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images. One that fits into cache easily and the other that does not fit into cache but fits into main memory. Both of the images were converted to single spectrum images represented by single precision numbers for each pixel. The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it. As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB. The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it. None of the multicore cache memories were big enough to store this image entirely.
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Results}
\label{sec:results}
\subsection{Performance of bilateral filtering kernel on CPU}
For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21. The spatial spread and the photometric spread were also set to 10. The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization. For Core i7 chip compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading, i.e., there are no more than one thread per core.

@@

\begin{wrapfigure}{r}{0.25\textwidth}
\centering
\includegraphics[width=0.25\textwidth]{images/nehalemtime}
\caption{Time taken by Intel Nehalem architecture}
\vspace{0mm}
\label{fig:nehalemtime}
\end{wrapfigure}
Intel Nehalem chip with 32 cores shows an impressive speedup as shown in Figure~\ref{fig:NehalemFinal}. The reason behind not showing a super-linear speedup in this case is the lack of enough parallelism that can be afforded by the kernel for this image. Figure~\ref{fig:nehalemtime} shows the absolute timings for the entire process, the timings tend to saturate with 32 processors as the total processing time approaches one second. Both optimizations still multiply their individual speedup to result in the combined speedup. The final speedup is 26.46x using 32 cores; the speedup only due to SSE was 1.12x and the speedup for BL\_Naive algorithm was 23.47x. 
\begin{figure*}[t!]
\begin{centering} 
\subfigure[Large 3000X3000 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10} \label{fig:hw10}} 
\subfigure[Large 3000X3000 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5} \label{fig:hw5} } 
\subfigure[Smaller 768X512 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
\subfigure[Smaller 768X512 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
\end{centering}
\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
\label{fig:gpuresults} 
\end{figure*}

@@

AMD Shanghai similarly shows an impressive speedup of 20.92x (Figure~\ref{fig:ShanghaiFinal}). The interesting case is with AMD Phenom as shown in Figure~\ref{fig:PhenomFinal}. The speedup drops a little as the number of cores increases beyond three. The reason behind this is the turbo core technology; when there are a maximum of three cores being used, the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.

\subsection{Performance of bilateral filtering kernel on GPU}

@@

To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds, while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.

>>>>

A previous study, closest to ours, is done by Xu et al.\cite{xu2009}, where they study the GPU-amenable implementation of bilateral filtering kernel. Our algorithms report 2x faster processing time than the timings reported in \cite{xu2009} on the same GPU - GTX 280.

In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time in half.
Our specific contributions are:
\begin{enumerate}
\item Improving the performance of bilateral filtering kernel on multicore machines by exploiting the availability of special purpose registers. These improvements resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\item Creating a novel pair-symmetric algorithm to reduce the number of calculations by a factor of two. 
\item Efficiently porting the pair-symmetric algorithm on GPU by exploiting the intra-warp synchronization. 
\end{enumerate}

@@

\begin{algorithm}[h]
\caption{Pair-symmetric Bilateral Filtering Algorithm}
\label{alg:Example-Algorithm-float-1}
\begin{algorithmic}
\STATE /* D is the set of dependency arrays */
\FORALL{element $s_{i,j}$ in source image s}
\STATE{$D_1[i,j] \leftarrow 0$}
\STATE{$D_2[i,j] \leftarrow 0$}
\ENDFOR
\FORALL{element $s_{i,j}$ in source image $s$}
\STATE{$X_1 \leftarrow 0$}
\STATE{$X_2 \leftarrow 0$}
\FOR{$k=1$ \TO $filter\_hw$}
\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
\STATE{/* $C_1$ and $C_2$ are user defined constants */}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
\ENDFOR
\STATE{syncthreads();}
\ENDFOR
\FOR{$l = -filter\_hw$ \TO $0$}
\STATE{$temp \leftarrow e^{C_{1} l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
\ENDFOR
\STATE $D_1[i,j] \leftarrow D_1[i,j] + X_1$
\STATE $D_2[i,j] \leftarrow D_2[i,j] + X_2$
\ENDFOR
\FORALL{element $s_{i,j}$ in source image s}
\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{table*}[t!]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache/Memory}  & Compiler\\
\hline 
Intel Xeon E5410  & 2.33 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1\\
\hline 
Intel Xeon X7560  & 2.27 $GH_z$  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1\\
\hline 
Intel Core i7-870  & 2.93 $GH_z$  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2\\
\hline 
AMD Opteron 2376  & 2.3 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1\\
\hline 
AMD Opteron 8350  & 2.0 $GH_z$  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 \\
\hline 
AMD Phenom II X6 1045T  & 2.7 $GH_z$  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2 \\
\hline 
NVidia GeForce GTX 280 & 1296 MHz & 240 & 1 & 240 & - & 1 GB GDDR3 & CUDA 2.0\\
\hline
\end{tabular}
\caption{Architectural details of chips employed for experiments}
\label{tab:archs} 
\end{table*}

@@

\textbf{Input Image:~}In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images. One that fits into cache easily and the other that does not fit into cache but fits into main memory. Both of the images were converted to single spectrum images represented by single precision numbers for each pixel. The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it. As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB. The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it. 
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Results}
\label{sec:results}
\subsection{Performance of bilateral filtering kernel on CPU}
For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21. The spatial spread and the photometric spread were also set to 10. The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization. For Core i7 chip compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading, i.e., there are no more than one thread per core.

@@

\begin{wrapfigure}{r}{0.25\textwidth}
\centering
\includegraphics[width=0.25\textwidth]{images/nehalemtime}
\caption{Time taken by Intel Nehalem architecture}
\vspace{0mm}
\label{fig:nehalemtime}
\end{wrapfigure}
Intel Nehalem chip with 32 cores shows an impressive speedup as shown in Figure~\ref{fig:NehalemFinal}. The reason behind not showing a super-linear speedup in this case is the lack of enough parallelism that can be afforded by the kernel for this image. Figure~\ref{fig:nehalemtime} shows the absolute timings for the entire process, the timings tend to saturate with 32 processors as the total processing time approaches one second. Both optimizations still multiply their individual speedup to result in the combined speedup. The final speedup is 26.46x using 32 cores; the speedup only due to SSE was 1.12x and the speedup for BL\_Naive algorithm was 23.47x. 
\begin{figure*}[t!]
\begin{centering} 
\subfigure[Large 3000X3000 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10} \label{fig:hw10}} 
\subfigure[Large 3000X3000 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5} \label{fig:hw5} } 
\subfigure[Smaller 768X512 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
\subfigure[Smaller 768X512 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
\end{centering}
\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card. X and Y represent the number of threads inside a block; number of blocks is listed next to the name of the algorithm (64x64 and 32x32).} 
\label{fig:gpuresults} 
\end{figure*}

@@

AMD Shanghai similarly shows an impressive speedup of 20.92x (Figure~\ref{fig:ShanghaiFinal}). The interesting case is with AMD Phenom as shown in Figure~\ref{fig:PhenomFinal}. The speedup drops a little as the number of cores increases beyond three. The reason behind this is the turbo core technology; when there are a maximum of three cores being used, the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Performance of bilateral filtering kernel on GPU}

@@

To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on Barcelona CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 131.98 seconds, while our pair-symmetric algorithm took 74.31 seconds - a 43.7\% reduction in execution time.
